---
title: "How To Learn Your Mate Preferences"
format: html
editor: visual
---

## Mate Preference Learning

**Purpose**: This model was designed to explore how might species-typical mating psychology give rise to individually tailored mate preferences. Using an agent-based model, we will test the ability of a reinforcement learning algorithm to reproduce the ideal mate preferences of N = 1,044 participants.

✅ **Agents**: Each agent represents a participant. The agents are made up of 15 traits, an initial preference vector (10's across all traits), an initial reward vector (date mate value given preferences), memory for reward vector, memory for preference vector.

**Life Cycle:**

1.  Mate choice for n cycles using preference vector.

    1.  Each cycle, they choose a mate that maximizes mv\*received/sent

    2.  Store mv\*received/sent in memory for reward vector (needs to be constantly recalc)

    3.  Preference vector goes into preference memory

2.  Change preferences relative to memory by methods:

    1.  Random

    2.  Decreased fixed amount (need to code diff for this)

    3.  Move towards highest reward partner (and this)

3.  Mate choice for n cycles using changed preference vector

    1.  Compare max reward to max reward stored in memory (do new preferences do better than old preferences?)

        1.  If better: Store new preferences in memory & store reward in memory

        2.  If not better: Keep memory unchanged; discard new preference & reward vector.

4.  Cycle back to #2 (change preferences relative to memory by methods---for now only Random) until preferences are no longer changing.

SIMULATION

Learning Cycles:

-   change preferences

-   calculate MV

-   create choice matrix

-   mate choice cycles

    -   calculate reward

    -   choose mate

    -   update choice matrix

-   compare reward

update as necessary

Take this models reward as is (before update preferences). Save to memory.

Copy and paste to for loop that has (new) ability to change preferences.

Random Update Methods

Check if this is just anchoring to mvs and randomly changing prefs to = mvs

Methods Cycle : randomMethod, absoluteMethod, etc.

Week 4 Have functional draft for a few methods

```{r}
#PARAMETERS

# the 15 Traits
traitNames <- paste0(c("affectionate", 
                "ambition", 
                "artistic", 
                "disposition", 
                "family", 
                "health", 
                "humor", 
                "intelligent", 
                "kind", 
                "parenting", 
                "physatt", 
                "religious", 
                "resources", 
                "sexy", 
                "status"))

# Starting Values for Prefernces, Reward Memory. 
startPref <- 10
startRMem <- 0

# Interest Ratio (starting values)
recieved <- 1
sent <- 1

# number of males and females
maleNum <- 100
femaleNum <-100

# Methods

methodList <- c("randomMethod", "absoluteMethod")


# absoluteMethod 
abValue <- .1


# Cycles 
methodCycles <- length(methodList)
prefLearningCycles <- 10000
prefTestingCycles <- 1000



```

Agent Generation

```{r}


# Human_data; PIN(3); sex(4); Ideal(235:249) Affectionate, Ambition, Artistic, Disposition, Family, Health, Humor, Intelligent, Kind, Parenting, Physical Attractiveness, Religious, Resources, Sexy, Status; Importance(283:297) " ; Self Composite(345:359) " ; Self Mate Value Composite(418) "

data <- read.csv("Human_data.csv")


data <- data[complete.cases(data[, c(235:249)]),]
# Sampled 10 people to check if things are working right. 

females <- data[data$sex == 0,]
males <- data[data$sex == 1,]


females <- females[sample(1:nrow(females), femaleNum),]
males <- males[sample(1:nrow(males), maleNum),]



# MALES: TRAITS ; PREFERENCE VECTOR, MEMEORY ; REWARD VECTOR, MEMORY
mTraits <- males[, c(345:359)]
mPref <- matrix(startPref, nrow(males), 15)
colnames(mPref) <- c("affectionate", 
                     "ambition", 
                     "artistic", 
                     "disposition", 
                     "family", 
                     "health", 
                     "humor", 
                     "intelligent", 
                     "kind", 
                     "parenting", 
                     "physatt", 
                     "religious", 
                     "resources", 
                     "sexy", 
                     "status")
mPrefMemory <- matrix(startPref, nrow(males), 15)
colnames(mPrefMemory) <- c("affectionate", 
                           "ambition", 
                           "artistic", 
                           "disposition", 
                           "family", 
                           "health", 
                           "humor", 
                           "intelligent", 
                           "kind", 
                           "parenting", 
                           "physatt", 
                           "religious", 
                           "resources", 
                           "sexy", 
                           "status")

mRewMemory <- matrix(startRMem, nrow(males), nrow(females))






# FEMALES: PREFERENCE VECTOR, MEMORY ; REWARD VECTOR, MEMORY
fTraits <- females[, c(345:359)]
fPref <- matrix(startPref, nrow(females), 15)
colnames(fPref) <- c("affectionate", 
                     "ambition", 
                     "artistic", 
                     "disposition", 
                     "family", 
                     "health", 
                     "humor", 
                     "intelligent", 
                     "kind", 
                     "parenting", 
                     "physatt", 
                     "religious", 
                     "resources", 
                     "sexy", 
                     "status")

fPrefMemory <- matrix(startPref, nrow(females), 15)
colnames(fPrefMemory) <-c("affectionate", 
                          "ambition", 
                          "artistic", 
                          "disposition", 
                          "family", 
                          "health", 
                          "humor", 
                          "intelligent", 
                          "kind", 
                          "parenting", 
                          "physatt", 
                          "religious", 
                          "resources", 
                          "sexy", 
                          "status")

fRewMemory <- matrix(startRMem, nrow(females), nrow(males))


# Putting the AGENTS together. Make traits, preferences, and preference memory as a part of the agent. The ENVIORNMENT is the rewards, reward memory, and the choices.
fAgents <- cbind(fTraits, fPref, fPrefMemory)
mAgents <- cbind(mTraits, mPref, mPrefMemory)





```

Functions

```{r}

# Mate Value Calculation : own pref vector, others trait composite
mvcalc <- function(pref, traits){
  names(pref) <- 1:15
  names(traits) <- 1:15
  mv <- 10 *(-1 *(dist(rbind(pref, traits))) + sqrt(10^2 *15)) / sqrt(10^2 *15)
  return(mv)
}

# Reward Calculation : interest recieved, interest sent
rcalc <- function(mv, recieved, sent) {
  
  interestRatio <- t(recieved) / sent
  
  # make interest ratio max = 1 , if recived / sent > 1 , set it to 1. # anti-stalker exploit 
  
  interestRatio[interestRatio > 1] <- 1
  
  reward <- mv * interestRatio 
  
  return(reward)
}

```

```{r}
# make randomMethod only take one prefernce at a time... on second thought, this could be controlled by parameter; randomly change n preferences at a time. 

# Random Method: Change row values for every col to a random number. Don't change highest reward one. 
# Base Random Method
randomMethod <- function(pref){
  
  random <- t(apply(pref, 1, function(y) round(runif(y, min = 0, max = 10), 1)))
  colnames(random) <- colnames(pref)
  
  changeMat <- cbind(1:nrow(pref), sample(1:15, nrow(pref), replace = T))
  pref[changeMat] <- random[changeMat]
  
  return(pref)
}
  
# How do I make this so it changes prefernces n at a time? And how do I make it so they change least if most important, and most if least important? 


absoluteMethod <- function(pref){
  
  # add a parameter where only n number get affected by absolute method (at random?)
  absolute <- pref - abValue 
  colnames(absolute) <- colnames(pref)
  
  changeMat <- cbind(1:nrow(pref), sample(1:15, nrow(pref), replace = T))
  pref[changeMat] <- absolute[changeMat]
  
  return(pref)
}




```

WORK IN PROGRESS

```{r}

mpUpdate <- function(method = "method", pref) {
  
  if(method == "randomMethod") {
    
    pref <- randomMethod(pref)
    
    } 
  
  if(method == "absoluteMethod") {
        
    pref <- absoluteMethod(pref)
    
    }
    
    
  return(pref)
}


```

# Simulation:

Does mate preference memory map onto own mate value somehow? IS this capturing \[... \] What do i think it should capture?

```{r}

 for(u in 1:methodCycles) {
   
# re-sets inital traits after a method; if not: old method pref updates will stay for next update method. I dont want that. 
   
   fPref <- matrix(startPref, nrow(females), 15)
   colnames(fPref) <- traitNames
   
   fPrefMemory <- matrix(startPref, nrow(females), 15)
   colnames(fPrefMemory) <- traitNames
   fRewMemory <- matrix(startRMem, nrow(females), nrow(males))
    
   
   mPref <- matrix(startPref, nrow(males), 15)
   colnames(mPref) <- traitNames
    
   mPrefMemory <- matrix(startPref, nrow(males), 15)
   colnames(mPrefMemory) <- traitNames
   mRewMemory <- matrix(startRMem, nrow(males), nrow(females))

   
   
   
# method cycle 
   
   method <- methodList[u]
   
   
  for(l in 1:prefLearningCycles) {
    
    # i think this is creating a problem: the if(l > 1) is offseting the u loop on and 
    if(l > 1){
      fPref <- mpUpdate(method, fPref)      
      mPref <- mpUpdate(method, mPref)
    }
    # Create choice matrix:
    fChoice <- matrix(1, nrow(females), nrow(males))
    mChoice <- matrix(1, nrow(males), nrow(females))
    
     # Calculate MV
    fMV <- t(apply(fPref, 1, function(x) apply(mTraits, 1, function (y) mvcalc(x, y))))
    mMV <- t(apply(mPref, 1, function(x) apply(fTraits, 1, function(y) mvcalc(x, y))))
    
    for(t in 1:prefTestingCycles){
      
      # Calculate Reward 
      fReward <- rcalc(fMV, mChoice, fChoice)
      mReward <- rcalc(mMV, fChoice, mChoice)
      
      
      # Choose Mate 
      fMaxReward <- cbind(1:nrow(females), apply(fReward, 1, which.max))
      mMaxReward <- cbind(1:nrow(males), apply(mReward, 1, which.max))
      
      
      
      
      # Update Choice Matrices
      fChoice[fMaxReward] <- fChoice[fMaxReward] + 1
      mChoice[mMaxReward] <- mChoice[mMaxReward] + 1
     
    
    }
    
# Commit to Memory 
# Store mReward & fReward in Reward Memory (of MaxReward mates?) and compare vector to memory
    
    for(f in 1:nrow(females)){
      maxNow <- max(fReward[f, ])
      maxMemory <- max(fRewMemory[f, ])
      
      if(maxNow >= maxMemory){
        fRewMemory[f, ] <- fReward[f, ]
        fPrefMemory[f, ] <- fPref[f, ]
      } else {
        fPref[f, ] <- fPrefMemory[f, ]  
        
        }
      }
    for(m in 1:nrow(males)){
      maxNow <- max(mReward[m, ])
      maxMemory <- max(mRewMemory[m, ])
        
      if(maxNow >= maxMemory){
        mRewMemory[m, ] <- mReward[m, ]
        mPrefMemory[m, ] <- mPref[m, ]
      } else{
        mPref[m, ] <- mPrefMemory[m, ]
        
      }
    } # male memory loop
    
    # save 2 diff things: calc average cor between predicted prefs and observed prefs. (to check in converging to value or just bouncing around) 
    # data.frame ^ 
    
    
# I need to make a for loop that creates a dataframe each loop. Then a for loop inside that to record (m/f)PrefMemory for each testing cycle. 
    
# NOPE ^ i need to create a dataframe every time it hits this part of the loop. Save the dataframe"t" <- #dataFromCurrentLoop 
    
    
    
    if(u == 1){
      
      assign(x = paste0("mRandPrefTest", t), value = mPrefMemory)
      assign(x = paste0("fRandPrefTest", t), value = fPrefMemory)
      
    } else if(u == 2){
      
      assign(x = paste0("mAbsPrefTest", t), value = mPrefMemory)
      assign(x = paste0("fAbsPrefTest", t), value = fPrefMemory)
      
    }
    
      
    
    
    
  } # prefLearningCycle loop
   
## Save runs to analyse ##
   
# Save randomMethods outputs; if u == 1, save (f/m)PrefMemory to data frame. 
   # Save end predicted preferences. 
   # data.frame^
   
# Save absoluteMethods outputs 
   if(u == 1){
     mRandLearned <- mPrefMemory
     fRandLearned <- fPrefMemory
     
   } else if(u == 2){
     
     mAbsLearned <- mPrefMemory
     fAbsLearned <- fPrefMemory
     
   }
   
} # methodCycles loop
 

# Testing Preferences Analysis (do values converge?)






# Learned Correlation 
mRandLearnedCor <- mean(diag(cor(mRandLearned, males[235:249])))
fRandLearnedCor <- mean(diag(cor(fRandLearned, females[235:249])))

mAbsLearnedCor <- mean(diag(cor(mAbsLearned, males[235:249])))
fAbsLearnedCor <- mean(diag(cor(fAbsLearned, females[235:249])))










  # side note, i can check how accurate my model works on single trait preferences. Have every prefernce besides one as the actual values. Then see if the model can accurately come up with a pref value that is similar to the actual value. 

```

First sentence ; organisms have preferences that guide them towards mates.
